Efficiently stable presentations from error-correcting codes

### 2023-10-13 efficient_stability.tex
intro & abstract edits

\mnote{Here I genuinely changed the phrasing. I don't think it is currently very good, but it is better than it was. Feel free to iterate:} Equipped with this tradeoff in mind, we analyze various natural examples of presentations. We provide a general method for constructing presentations of $\Z_2^k$ from linear error-correcting codes. We observe that the resulting presentation has a weak form of stability exactly when the code is  \emph{locally testable}.\mnote{I have a problem with the notion of locally testable. The notion reffers ususally to testable LDPCs. We do not need the low density to define our notions and construction, only the testable part. I think we should change it to just testable codes... What do you think? This appears throughout the intro... It won't be a major change} This raises the question of whether locally testable codes give rise to genuinely stable presentations using this method.  While we cannot show that in general, we leverage recent results in the study of non-local games in quantum information theory (Ji et al., Discrete Analysis 2021) to show that a specific instantiation of our construction, based on the Reed-Muller family of codes, leads to a stable presentation of $\Z_2^k$ of size $\poly\log(k)$ only. As an application, we combine this result with recent work of de la Salle (arXiv:2204.07084) to re-derive  the quantum low-degree test of Natarajan and Vidick (IEEE FOCS'18), which is a key building block in the recent refutation of Connes' Embedding Problem via complexity theory (Ji et al., arXiv:2001.04383). 
\end{abstract}

@@

We make an observation that connects the study of locally testable codes to questions of stability in group theory and motivates our work. Let $\code=\ker h$ be a linear code as above, and suppose for simplicity that $\F=\F_2$ is the binary field. Consider the finitely presented group 
\begin{align}
 G(h) \,=\, \langle S:R\rangle \,=\, \big\langle &x_1,\ldots,x_n \,:\quad  x_j^2=e\quad \forall 1\leq j\leq n\;,\notag\\
& \quad \prod_{1\leq j \leq n} x_j^{h_{ij}} = e\;,\quad [x_{j_1},x_{j_2}]^{h_{ij_1} h_{ij_2}}=e\quad \forall 1\leq i\leq m\,\forall 1\leq j_1,j_2\leq n\,\big\rangle\;.\label{eq:gh-intro}
\end{align}
Here a commutation relation between two generators  $[x_{j_1},x_{j_2}]=e$ is imposed only when needed for the relations $\prod_{1\leq j \leq n} x_j^{h_{ij}} = e$ to make sense, i.e.\ two generators are required to commute only if they both take part in the same equation, which is the case if and only if $h_{ij_1}h_{ij_2}=1$ for some $i$.

@@

\mnote{The following two paragraphs are problematic. The size of $G(h)$ is independent of $n,k$ in general, even if it is finite. The only case in which you can bound the size of $G(h)$ according to $n,k$ is when it is \textbf{abelian}, and we insisted on not assuming that. I feel like we should add all commutation relations before discussing this part.} However the results of~\cite{gowers2017inverse,de2019operator} do not imply the same quantitative stability bounds for $G(h)$ with respect to its defining presentation~\eqref{eq:gh-intro} (even if $G(h)$ is given to be finite). The reason to prefer the presentation~\eqref{eq:gh-intro} as opposed to the multiplication table presentation of $G(h)$ is that~\eqref{eq:gh-intro} is much more succinct: if $n,k$ are linearly related then it has $\textrm{poly}(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Such a gain is essential when one recalls the interpretation of the presentation $G(h)$ as a ``tester'' for $G$ --- the size of the presentation is then directly related to the amount of \emph{randomness} required by the tester (to sample a random relation); and in computer science applications randomness is seen as an essential resource (we discuss this more below, in the context of quantum computing).

@@

To summarize, approximate stability results for $G(h)$ enable one to obtain structural results about near-optimal strategies in $\game_h$ --- such a result is known as a \emph{rigidity} result in quantum information. A rigidity result about a game called the quantum low-degree test~\cite{natarajan2018low} is at the heart of the proof of $\sf MIP^* = \sf RE$. The analysis of this test requires an efficient stability result for the Weyl-Heisenberg group. Informally, the reason that the stability result needs to be for an efficient presentation of the Weyl-Heisenberg group \mnote{We mix Weyl-Heisenberg with Pauli all the time. In the main text we call them Pauli, and mention that they are a special case of a Heisenberg group. I like the Weyl-Heisenberg name, since we ignore the y operators in this case, and I thought they are formally part of the Pauli group. But, since we already use notations with P throughout, it would be probably easier to change to Pauli here as well. }, as opposed to e.g.\ the multiplication table presentation, is because to obtain the final result it is necessary that the ``complexity'' of the test, or game, is smaller than the ``complexity'' of the object, or group being tested. Of course here we are loose about what we mean by ``complexity,'' and refer to~\cite{ji2021mip,vidickmip} for high-level explanations. The quantum low-degree test is described and analyzed in Section~\ref{sec:pbt}. We end by mentioning an open question: if one was able to obtain $|S_k|,|R_k|=\poly(k)$ while also having $\delta(\eps)=\poly(\eps)$ in the informal theorem stated above, then this result would likely, through the connection we just described, have consequences for the efficient verification of quantum computations in the framework of interactive proof systems---see e.g.~\cite{coladangelo2019verifier,natarajan2023bounding} for a sample of known results in this direction.

@@

\tnote{more? open questions?}\mnote{If we have open problems, maybe we should collect them at the end? Not sure about that. Did you think of any open problems other than the permutation analogue of the quantum soundness of the low-deg test?}

@@

\subsection{Braiding the code test}

>>>>

\mnote{Here I genuinely changed the phrasing. I don't think it is currently very good, but it is better than it was. Feel free to iterate:}\tnote{slightly edited}With this goal in mind we analyze various natural examples of presentations. We provide a general method for constructing presentations of $\Z_2^k$ from linear error-correcting codes. We observe that the resulting presentation has a weak form of stability exactly when the code is  \emph{locally testable}.\mnote{I have a problem with the notion of locally testable. The notion reffers ususally to testable LDPCs. We do not need the low density to define our notions and construction, only the testable part. I think we should change it to just testable codes... What do you think? This appears throughout the intro... It won't be a major change}\tnote{Right, it's not clear what motivation there is for the ``locally'' aspect in the language of groups; that the relations be ``short''---is that ever relevant? I'm ok for dropping ``locally''. It's not hard as we can search for the word.} This raises the question of whether locally testable codes give rise to genuinely stable presentations using this method.  While we cannot show that this is the case in general, we leverage recent results in the study of non-local games in quantum information theory (Ji et al., Discrete Analysis 2021) to show that a specific instantiation of our construction, based on the Reed-Muller family of codes, leads to a stable presentation of $\Z_2^k$ of size $\poly\log(k)$ only. As an application, we combine this result with recent work of de la Salle (arXiv:2204.07084) to re-derive  the quantum low-degree test of Natarajan and Vidick (IEEE FOCS'18), which is a key building block in the recent refutation of Connes' Embedding Problem via complexity theory (Ji et al., arXiv:2001.04383). 
\end{abstract}

@@

We make an observation that connects the study of locally testable codes to questions of stability in group theory and motivates our work. Let $\code=\ker h$ be a linear code as above, and suppose for simplicity that $\F=\F_2$ is the binary field. Consider the finitely presented group 
\begin{align}
 G(h) \,=\, \langle S:R\rangle \,=\, \big\langle &x_1,\ldots,x_n \,:\quad  x_j^2=e\quad \forall 1\leq j\leq n\;,\notag\\
& \quad \prod_{1\leq j \leq n} x_j^{h_{ij}} = e\;,\quad [x_{j_1},x_{j_2}]^{h_{ij_1} h_{ij_2}}=e\quad \forall 1\leq i\leq m\,\forall 1\leq j_1,j_2\leq n\,\big\rangle\;.\label{eq:gh-intro}
\end{align}
Here a commutation relation between two generators  $[x_{j_1},x_{j_2}]=e$ is imposed only when needed for the relations $\prod_{1\leq j \leq n} x_j^{h_{ij}} = e$ to make sense, i.e.\ two generators are required to commute only if they both take part in the same equation, which is the case if and only if $h_{ij_1}h_{ij_2}=1$ for some $i$. \tnote{added:}While one could add all pairwise commutation relations, forcing $G(h)$ to be Abelian --- and indeed we will do this in some cases later --- we choose the specific presentation~\eqref{eq:gh-intro} for consistency with the literature on non-local games, and particularly so-called \emph{linear constraint system games}, which we review in more detail below.

@@

\tnote{modified from here:}
To simplify the problem let us consider the following presentation for an obviously finita and Abelian group
\begin{align}
  \widetilde{G(h)} \,=\, \langle S:R\rangle \,=\, \big\langle &x_1,\ldots,x_n \,:\quad  x_j^2=e\quad \forall 1\leq j\leq n\;,\notag\\
 & \quad \prod_{1\leq j \leq n} x_j^{h_{ij}} = e\;,\quad [x_{j_1},x_{j_2}]=e\quad \forall 1\leq i\leq m\,\forall 1\leq j_1,j_2\leq n\,\big\rangle\;.\label{eq:wtgh-intro}
 \end{align}
This is the same as $G(h)$, except that all pairwise commutations have been added. As we show formally later (see Lemma~\ref{lem:com-code}), it is not hard to check that $\widetilde{G(h)}$ is a presentation of the group $\Z_2^k$, where recall that $k=\dim\ker h$. Our most important results pertain to the stability of the presentation $\widetilde{G(h)}$; although we will later (Section~\ref{sec:braiding}) consider some non-Abelian extensions of it.

\newcommand{\wtG}{\widetilde{G(h)}}

\mnote{The following two paragraphs are problematic. The size of $G(h)$ is independent of $n,k$ in general, even if it is finite. The only case in which you can bound the size of $G(h)$ according to $n,k$ is when it is \textbf{abelian}, and we insisted on not assuming that. I feel like we should add all commutation relations before discussing this part.}\tnote{Yes so I tried doing this above, is it better? If we keep the notation, we need to modify around Lemma~\ref{lem:com-code} as well}

Importantly for us, the results of~\cite{gowers2017inverse,de2019operator} do not imply the same quantitative stability bounds for $\wtG$ with respect to its defining presentation~\eqref{eq:wtgh-intro}. The reason to prefer the presentation~\eqref{eq:wtgh-intro} as opposed to the multiplication table presentation of $\wtG$ is that~\eqref{eq:wtgh-intro} is much more succinct: if $n,k$ are linearly related then it has $\textrm{poly}(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Such a gain is essential when one recalls the interpretation of the presentation $\wtG$ as a ``tester'' for $G$ --- the size of the presentation is then directly related to the amount of \emph{randomness} required by the tester (to sample a random relation); and in computer science applications randomness is seen as an essential resource (we discuss this more below, in the context of quantum computing).

@@

To summarize, approximate stability results for $G(h)$ enable one to obtain structural results about near-optimal strategies in $\game_h$ --- such a result is known as a \emph{rigidity} result in quantum information. A rigidity result about a game called the quantum low-degree test~\cite{natarajan2018low} is at the heart of the proof of $\sf MIP^* = \sf RE$. The analysis of this test requires an efficient stability result for the Weyl-Heisenberg group. Informally, the reason that the stability result needs to be for an efficient presentation of the Weyl-Heisenberg group \mnote{We mix Weyl-Heisenberg with Pauli all the time. In the main text we call them Pauli, and mention that they are a special case of a Heisenberg group. I like the Weyl-Heisenberg name, since we ignore the y operators in this case, and I thought they are formally part of the Pauli group. But, since we already use notations with P throughout, it would be probably easier to change to Pauli here as well. }\tnote{Well the P is a macro, and sometimes we write ``for all $W\in\{X,Z\}$'', so if you think that WH would be more appropriate I think that we could do it}, as opposed to e.g.\ the multiplication table presentation, is because to obtain the final result it is necessary that the ``complexity'' of the test, or game, is smaller than the ``complexity'' of the object, or group being tested. Of course here we are loose about what we mean by ``complexity,'' and refer to~\cite{ji2021mip,vidickmip} for high-level explanations. The quantum low-degree test is described and analyzed in Section~\ref{sec:pbt}. We end by mentioning an open question: if one was able to obtain $|S_k|,|R_k|=\poly(k)$ while also having $\delta(\eps)=\poly(\eps)$ in the informal theorem stated above, then this result would likely, through the connection we just described, have consequences for the efficient verification of quantum computations in the framework of interactive proof systems---see e.g.~\cite{coladangelo2019verifier,natarajan2023bounding} for a sample of known results in this direction.

@@

\tnote{more? open questions?}\mnote{If we have open problems, maybe we should collect them at the end? Not sure about that. Did you think of any open problems other than the permutation analogue of the quantum soundness of the low-deg test?}\tnote{There's the obvious open problem of finding an LTC that is not stable (since I'd conjecture it this way), and conversely other examples of codes that are stable. We can also ask how far things can be pushed; in principle we could have stability, in our sense, for a finite pres of an infinite group.}

@@

\subsection{Braiding the code test}
\label{sec:braiding}
